Structure Learning for $\VAR(p)$ Models

### 2019-05-01 paper/paper.tex
Moved paper to subfolder and added some more remarks.

\def\gc{\overset{\text{GC}}{\rightarrow}}  % Granger causality arrow
\def\pwgc{\overset{\text{PW}}{\rightarrow}}  % Granger causality arrow
\def\te{\overset{\mathcal{T}}{\rightarrow}}  % Transfer entropy arrow
\def\gcg{\mathcal{G}}  % Granger-causality graph
\def\VAR{\mathsf{VAR}}  % VAR(p) model

\newcommand{\linE}[2]{\hat{\E}[#1\ |\ #2]}  % Linear projection
\newcommand{\linEerr}[2]{\xi[#1\ |\ #2]}  % Error of linear projection
\newcommand{\pa}[1]{P(#1)}  % Parents of a node
\newcommand{\anc}[1]{\mathcal{A}(#1)}  % Ancestors of a node

@@

\graphicspath{{./figures/}}

\title{Structure Learning for $VAR(p)$ Models}
\author{R. J. Kinnear, R. R. Mazumdar}

@@

\begin{theorem}[Granger Causality Equivalences]
  Let $x(t)$ be a WSS process with absolutely summable covariance
  sequence, and uniformly bounded spectral density.  Denote by
  $\xi_{ij} \defeq \linEerr{x_i(t)}{\mathcal{H}^{-j}}$ and
  $\xi_i \defeq \linEerr{x_i(t)}{\mathcal{H}_t}$.  Then, the following are equivalent:

  \begin{enumerate}
    \item{$x_j \not\gc x_i$}
    \item{$\forall \tau \in \N_+\ B_{ij}(\tau) = 0$}
    \item{$\mathcal{F}_{j \rightarrow i} \defeq \ln\frac{\xi_i}{\xi_{ij}} = 0$}
    \item{$\mathcal{H}_t^{i} \perp \mathcal{H}_{t - 1}^{j}\ |\ \mathcal{H}_{t - 1}^{-j}$}
    \item{\hl{Wold $A(\tau)$ condition}}
    \item{Tie in the 0s of $S(\omega)^{-1}$, though this isn't immediately directed.}
  \end{enumerate}
\end{theorem}

The equivalence $(1) \iff (3)$ is essentially a restatement of the
definition, but is in line with the seminal work of Geweke
\cite{geweke1982}, \cite{geweke1984}.  Statement (4) means that
$(y - \linE{y}{\mathcal{H}_{t - 1}^{-j}}) \perp (z -
\linE{z}{\mathcal{H}_{t - 1}^{-j}})\ \forall y \in \mathcal{H}_t^i,\ z \in
\mathcal{H}_{t - 1}^j$, which is an appealing geometric statement that
there is no direct feedback from $x_j$ to $x_i$ \cite{lindquist}.

\begin{proof}
$(1) \implies (2)$: The projection for $x_i(t)$ onto $\mathcal{H}_{t - 1}$ is (by definition) given by a form similar to equation \ref{eqn:ar_representation}:

\[
  \E|x_i(t) - \linE{x_i(t)}{\mathcal{H}_{t - 1}}|^2 = \E|v_i(t) + \sum_{\tau = 1}^\infty \sum_{k = 1}^n (B_{i, k}(\tau) - \hat{B}_{i, k}(\tau))x_k(t - \tau)|^2.
\]

 Since $v(t)$ in equation \ref{eqn:ar_representation} is temporally uncorrelated, it follows that the optimal projection is given by the model coefficients themselves.  This holds similarly for the projection onto $\mathcal{H}_{t - 1}^{-j}$.  Then by $(1)$ we have

@@

$(2) \implies (4)$: In computing $(y - \linE{y}{\mathcal{H}_{t = 1}^{-j}})$ for $y \in \mathcal{H}_t^i$ it is sufficient to consider $y = x_i(t)$ since $\mathcal{H}_{t - 1}^i \subseteq \mathcal{H}_{t - 1}^{-j}$, in which case $(x_i(t) - \linE{x_i(t)}{\mathcal{H}_{t = 1}^{-j}}) = v_i(t)$.  $(4)$ follows since $v_i(t) \perp \mathcal{H}_{t - 1}$ and $\forall z \in \mathcal{H}_{t - 1}^j\ (z - \linE{z}{\mathcal{H}_{t - 1}^{-j}}) \in \mathcal{H}_{t - 1}$

@@

The following propositions justify various modifications of $x(t)$
applied in practice to massage $x(t)$ into a form amenable to more
standard tools.

\begin{theorem}[General Invariance]
  \hl{Under what conditions is this actually true?}

  Let $\zeta(t)$ be a stationary discrete time stochastic process.  Let
  $F, G$ be (possibly nonlinear and time varying) invertible filtering
  operations and $f(t)$, $g(t)$ be perfectly predictable and such that
  $x_j(t) \defeq F(\zeta_j - f)(t), x_i(t) \defeq G(\zeta_i - g)(t)$ are W.S.S.  Then,

\begin{equation}
    x_j \gc x_i \iff \zeta_j \te \zeta_i
  \end{equation}
\end{theorem}

\begin{corollary}[Invariance Under Invertible and Deterministic Modifications]
  Let $F(z), G(z)$ describe univariate and invertible
  linear-time-invariant filters.  And, let $f(t), g(t)$ be perfectly
  predictable processes.  Then,

  \begin{equation}
    x_j \gc x_i \iff F(x_j - f)(t) \gc G(x_i - g)(t)
  \end{equation}
\end{corollary}

\subsection{Pairwise Granger Causality}
Recall that Granger-causality in general must be understood with
respect to a particular universe of observations.  If $x_j \gc x_i$
with respect to $x_{-k}$, it may not hold with respect to $x$.  That
is, $x_k$ may be a common ancestor that when observed completely
explains the connection from $x_j$ to $x_i$.  In particular, we will
study \textit{pairwise} Granger-causality, where $x_i \pwgc x_j$ if
$x_i$ Granger-causes $x_j$ with respect only to $(x_i, x_j)$.

We establish first some notation.  Remaining in the context of a
collection of $n$ WSS processes
$x(t) = \big(x_1(t), \ldots, x_n(t)\big)$, we define the Granger-causality
graph $\gcg$ to be the directed graph formed on $n$ vertices where an
edge $(j, i) \in \gcg$ if and only if $x_j \gc x_i$.  We refer to the
set of parents of $x_i$ in $\gcg$ by $\pa{i}$, that is,
$j \in \pa{i} \iff x_j \gc x_i$.  The set of \textit{ancestors} of
$x_i$ in $\gcg$ will be denoted $\anc{i}$, that is, $j \in \anc{i}$ if
and only if there is a directed path in $\gcg$ from $j$ to $i$.  Even
though the definitions of Granger-causality naturally suggests that
each node in the graph should have a self-loop (unless the node is
simply independent noise) we will not include nodes in their own
parent sets: $i \not\in \pa{i} \subseteq \anc{i}$.  We note that $\gcg$ need not be a DAG.

The edges of the graph can be given a general notion of ``weight'' by
associating an edge $(j, i)$ with an LTI filter
$\{B_{ij}(\tau)\}_{\tau = 1}^\infty$ from the $\VAR$ representation of equation
\ref{eqn:ar_representation}.

% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
% I'm actually not sure this is true, I think that if there is a common ancestor it is
% ambiguous.  This may be true for transfer entropy, but for Granger-causality even if
% the information arriving at i is later than to j, it may be less corrupt and hence
% still facilitate better predictions about k's future.

Finally, the \textit{propagation delay} from $j$ to $i$, denoted $d(j, i)$, will be defined as

\[
  d(j, i) = \text{min}\big\{\sum_{k = 0}^{K - 1} d(p_k, p_{k + 1}})\ |\ i = p_0 \rightarrow p_1 \rightarrow \cdots \rightarrow p_K = j \text{ is a path in } \gcg \big\}
\]

and $d(j, i) \defeq \text{min}\{B_{i, j}(\tau) \ne 0\}$ if $j \in \pa{i}$.

\begin{lemma}
  \label{lem:pwgc_anc}
  $x_j \pwgc x_i$ if and only if at least one of the following hold

  \begin{enumerate}
    \item{$j \in \anc{i}$}
    \item{$\anc{i}\cap\anc{j} \ne \emptyset$.} %and $d(k, j) < d(k, i)$}
  \end{enumerate}
\end{lemma}

% \begin{remark}
%   This lemma formalizes the intuition of \textit{information flow} in $\gcg$.  We have $x_j \pwgc x_i$ whenever $i$ is descended from $j$, or if they have a common ancestor (call it $k$) and ``information'' from $k$ reaches $j$ before it reaches $i$.
% \end{remark}

\begin{proof}
\end{proof}

% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -

\begin{theorem}[Tree Recovery]
  \label{thm:tree_recovery}
  If the Granger-Causality graph $\gcg$ for $x(t)$ is a
  directed forest, then $\gcg$ can be inferred from pairwise
  causality tests alone.
\end{theorem}

\section{Structure Learning}
\subsection{Local Search Heuristics}

>>>>

\def\gc{\overset{\text{GC}}{\rightarrow}}  % Granger causality arrow
\def\pwgc{\overset{\text{PW}}{\rightarrow}}  % Granger causality arrow
\def\te{\overset{\mathcal{T}}{\rightarrow}}  % Transfer entropy arrow
\def\gcg{\mathcal{G}}  % Granger-causality graph
\def\VAR{\mathsf{VAR}}  % VAR(p) model
\def\B{\mathsf{B}}  % Filter B
\def\wtB{\widetilde{\B}}  % General filter B
\def\A{\mathsf{A}}  % Filter A
\def\H{\mathcal{H}}  % Hilbert space

\newcommand{\linE}[2]{\hat{\E}[#1\ |\ #2]}  % Linear projection
\newcommand{\linEerr}[2]{\xi[#1\ |\ #2]}  % Error of linear projection
\newcommand{\pa}[1]{pa(#1)}  % Parents of a node
\newcommand{\anc}[1]{\mathcal{A}(#1)}  % Ancestors of a node
\newcommand{\ancn}[2]{\mathcal{A}_{#1}(#2)}  % nth ancestors of a node
\newcommand{\gpn}[2]{gp_{#1}(#2)}  % nth generation grandparents
\newcommand{\wtalpha}[2]{\widetilde{\alpha}(#1, #2)}  % Some notation for lem:pwgc_anc
\newcommand{\dist}[2]{\mathsf{d}(#1, #2)}  % Distance between things
\newcommand{\gcgpath}[2]{#1 \rightarrow \cdots \rightarrow #2}  % A shorter path command

@@

\graphicspath{{../figures/}}

\title{Structure Learning for $\VAR(p)$ Models}
\author{R. J. Kinnear, R. R. Mazumdar}

@@

The equations (\ref{eqn:wold}), (\ref{eqn:ar_representation}) can be
represented as $x(t) = \A(z)w(t) = \B(z)x(t) + v(t)$ via the action of
the operators (LTI filters)
$\A(z) \defeq \sum_{\tau = 0}^\infty A(\tau)z^{-\tau}$ and
$\B(z) \defeq \sum_{\tau = 1}^\infty B(\tau)z^{-\tau}$ where the
operator $z^{-1}$ is the backshift operator acting on $\H$.  Since
$||z^{-1}|| = 1$, the inversion formulae

\begin{equation}
  \label{eqn:lsi_inversion}
  \A(z) = (I - \B(z))^{-1} = \sum_{k = 0}^\infty \B(z)^k
\end{equation}

are valid.
\subsection{Granger Causality}

\begin{definition}[Granger Causality]
  For the WSS series $x(t)$ we say that component $x_j$
  \textit{Granger-Causes} (GC) component $x_i$ (with respect to $x$)
  and write $x_j \gc x_i$ if given Hilbert spaces
  $\H_{t - 1}$, $\H^{-j}_{t - 1}$

\begin{equation}
  \linEerr{x_i(t)}{\H_{t - 1}} < \linEerr{x_i(t)}{\H^{-j}_{t - 1}},
\end{equation}

where $\xi[x \ |\ \H] = \E (x - \linE{x}{\H})^2$ is the mean squared
estimation error and $\linE{x}{\H} = \text{proj}_{\H}(x)$ denotes the
(unique) projection onto the Hilbert space $\H$.
\end{definition}

This notion captures the idea that the process $x_j$ provides
information about $x_i$ that is not available from elsewhere.  The
caveat ``with respect to $x$'' is important in that GC relations can
change when components are added to or removed from our collection $x$
of observations, e.g. new GC relations can arise if we remove the
observations of a common cause, and existing GC relations can
disappear if we observe a new mediating series.

The notion is closely related to the information theoretic measure of
transfer entropy, indeed, if everything is Gaussian then they are
equivalent \cite{barnett2009granger}.

\begin{theorem}[Granger Causality Equivalences]
  \label{thm:granger_causality_equivalences}
  Let $x(t)$ be a WSS process with absolutely summable covariance
  sequence, and spectral density uniformly bounded above and below.  Denote by
  $\xi_{ij} \defeq \linEerr{x_i(t)}{\H^{-j}}$ and
  $\xi_i \defeq \linEerr{x_i(t)}{\H_t}$.  Then, the following are equivalent:

  \begin{enumerate}
    \item{$x_j \not\gc x_i$}
    \item{$\forall \tau \in \N_+\ B_{ij}(\tau) = 0$}
    \item{$\mathcal{F}_{j \rightarrow i} \defeq \ln\frac{\xi_i}{\xi_{ij}} = 0$}
    \item{$\H_t^{i} \perp \H_{t - 1}^{j}\ |\ \H_{t - 1}^{-j}$}
    \item{$\linE{x_i(t)}{\H_{t - 1}^{-j}} = \linE{x_i(t)}{\H_{t - 1}}$}
    % \item{\hl{Wold $A(\tau)$ condition}}
    % \item{Tie in the 0s of $S(\omega)^{-1}$, though this isn't immediately directed.}
  \end{enumerate}
\end{theorem}

Statement (4) means that
$(y - \linE{y}{\H_{t - 1}^{-j}}) \perp (z - \linE{z}{\H_{t -
    1}^{-j}})\ \forall y \in \H_t^i,\ z \in \H_{t - 1}^j$, which is an
appealing geometric statement that there is no direct feedback from
$x_j$ to $x_i$ \cite{lindquist}.

\begin{proof}
  The equivalence $(1) \iff (3)$ is essentially a restatement of the
  definition, but is in line with the seminal work of Geweke
  \cite{geweke1982measurement}, \cite{geweke1984}.

  The equivalence $(3) \iff (5)$ is also immediate from the
  definitions i.e. the projections are equal if and only if the errors
  are equal.

  $(1) \implies (2)$: The projection for $x_i(t)$ onto $\H_{t - 1}$ is
  (by definition) given by a form similar to equation
  \ref{eqn:ar_representation}:

\[
  \E|x_i(t) - \linE{x_i(t)}{\H_{t - 1}}|^2 = \E|v_i(t) + \sum_{\tau = 1}^\infty \sum_{k = 1}^n (B_{i, k}(\tau) - \hat{B}_{i, k}(\tau))x_k(t - \tau)|^2.
\]

Since $v(t)$ in equation \ref{eqn:ar_representation} is temporally
uncorrelated, it follows that the optimal projection is given by the
model coefficients themselves.  This holds similarly for the
projection onto $\H_{t - 1}^{-j}$.  Then by $(1)$ we have

@@

$(2) \implies (4)$: In computing $(y - \linE{y}{\H_{t = 1}^{-j}})$ for
$y \in \H_t^i$ it is sufficient to consider $y = x_i(t)$ since
$\H_{t - 1}^i \subseteq \H_{t - 1}^{-j}$, in which case
$(x_i(t) - \linE{x_i(t)}{\H_{t = 1}^{-j}}) = v_i(t)$.  $(4)$ follows
since $v_i(t) \perp \H_{t - 1}$ and
$\forall z \in \H_{t - 1}^j\ (z - \linE{z}{\H_{t - 1}^{-j}}) \in \H_{t
  - 1}$

\hl{TODO: $(4) \implies (1)$}

@@

% The following propositions justify various modifications of $x(t)$
% applied in practice to massage $x(t)$ into a form amenable to more
% standard tools.

% \begin{theorem}[General Invariance]
%   \hl{Under what conditions is this actually true?}

%   Let $\zeta(t)$ be a stationary discrete time stochastic process.  Let
%   $F, G$ be (possibly nonlinear and time varying) invertible filtering
%   operations and $f(t)$, $g(t)$ be perfectly predictable and such that
%   $x_j(t) \defeq F(\zeta_j - f)(t), x_i(t) \defeq G(\zeta_i - g)(t)$ are W.S.S.  Then,

% \begin{equation}
%     x_j \gc x_i \iff \zeta_j \te \zeta_i
%   \end{equation}
% \end{theorem}

\begin{proposition}[Invariance Under Invertible and Deterministic Modifications]
  Let $x(t)$ be a WSS process with absolutely summable covariance
  sequence, and spectral density uniformly bounded above and below.  Let
  $F(z), G$ represent univariate and invertible
  linear-time-invariant filters.  And, let $f(t), g(t)$ be perfectly
  predictable processes.  Then,

  \begin{equation}
    x_j \gc x_i \iff F(x_j - f)(t) \gc G(x_i - g)(t)
  \end{equation}

\begin{proof}
  \hl{TODO}
\end{proof}
\end{proposition}

\subsection{Pairwise Granger Causality}
\label{sec:pwgc}
Recall that Granger-causality in general must be understood with
respect to a particular universe of observations.  If $x_j \gc x_i$
with respect to $x_{-k}$, it may not hold with respect to $x$.  For
example, $x_k$ may be a common ancestor which when observed completely
explains the connection from $x_j$ to $x_i$.  In particular, we will
study \textit{pairwise} Granger-causality.

\begin{definition}[Pairwise Granger-causality]
  We will say that $x_j$ pairwise Granger-causes $x_i$ and write
  $x_j \pwgc x_i$ if $x_i$ Granger-causes $x_j$ with respect only to
  $(x_i, x_j)$.
\end{definition}

This notion is of interest for a variety of reasons.  From a purely
theoretical or conceptual standpoint, we will seek to understand when
and how pairwise causality can capture the idea of ``flow of
information'' in the underlying graph.  It will also be useful to
reason about what the conditions under which \textit{unobserved}
components of $x(t)$ may or may not interfere with inference in the
actually observed components.  Finally, we will seek to construct
practical estimation procedures based purely on pairwise causality
tests and understand their properties.

We first need to establish some graph theoretic notation and
terminology, collected formally in definitions for the reader's
convenient reference.

\begin{definition}[Causality graph, Parents]

Remaining in the context of a collection of $n$ WSS processes
$x(t) = \big(x_1(t), \ldots, x_n(t)\big)$, we define the Granger-causality
graph $\gcg$ to be the directed graph formed on $n$ vertices where an
edge $(j, i) \in \gcg$ if and only if $x_j \gc x_i$.

We refer to the set of \textit{parents} of $x_i$ in $\gcg$ by
$\pa{i}$, that is, $j \in \pa{i} \iff x_j \gc x_i$.
\end{definition}

\begin{definition}[Grandparents, ancestors]
The set of level $\ell$ \textit{grandparents} of node $i$, denoted
$\gpn{\ell}{i}$, is the set such that $j \in \gpn{\ell}{i}$ if and only if
there is a \textit{directed and cycle-free path} of length $\ell$ in
$\gcg$ from $j$ to $i$ \textit{or} if $j = i$ and there is a cycle of
length $\ell$ starting at $i$ and ending back at $i$, that is, $i$ can be
it's own grandparent.  Clearly, $\pa{i} = \gpn{1}{i}$.

Finally, the set of \textit{level $\ell$ ancestors} of
$i$: $\ancn{\ell}{i} = \bigcup_{\lambda \le \ell}\gpn{\lambda}{i}$ is the set such that
$j \in \ancn{\ell}{i}$ if and only if there is a directed and cycle-free
path of length $\ell$ \textit{or less} in $\gcg$ from $j$ to $i$
\textit{or} if $j = i$ and there is a directed cycle of length $\ell$ or
less from $i$ back to $i$ in $\gcg$.  The set of \textit{all
  ancestors} of $i$ (i.e. $\ancn{n}{i}$) is denoted simply $\anc{i}$.
\end{definition}

Even though the definitions of Granger-causality naturally suggests
that each node in the graph should have a self-loop (unless the node
is simply independent noise) we will not include nodes in their own
parent sets: $i \not\in \pa{i} \subseteq \anc{i}$.  However, we do not
immediately assume that $\gcg$ is a DAG, and so \textit{we will allow for $i$
to be it's own ancestor}.

The edges of the Granger-causality graph $\gcg$ can be given a general
notion of ``weight'' by associating an edge $(j, i)$ with a
\textit{strictly causal} LTI filter

\begin{equation}
  \label{eqn:filter}
  \B_{ij}(z) = \sum_{\tau = 1}^{\infty} B_{ij}(\tau)z^{-\tau}
\end{equation}

from the $\VAR$ representation of equation
\ref{eqn:ar_representation}.  We will represent the action
(convolution) of this filter compactly by taking $z^{-1}$ as the
backshift operator on $\ell_2^n(\Omega, \mathcal{F}, \mathbb{P})$ when
convenient

\begin{equation}
  \sum_{\tau = 1}^\infty B_{ij}(\tau)x_j(t - \tau) \defeq \B_{ij}(z)x_j(t).
\end{equation}

% What if we take this as the definition of PWGC?  then, in a simply causal graph
% the "normal" definition becomes a theorem.  Or, if lem:pwgc_anc is true then
% maybe that is good enough?

From the $\VAR$ representation of $x(t)$ there is clearly a tight
relationship between each node and it's parent nodes, indeed

\begin{equation}
  \label{eqn:parent_expansion}
  x_i(t) = v_i(t) + \B_{ii}(z)x_i(t) + \sum_{k \in \pa{i}}\B_{ik}(z)x_k(t).
\end{equation}

% When is this guaranteed w.r.t. the full system?
Furthermore, if the filter $\B_{ii}$ is invertible, we can write

\begin{equation*}
  \begin{aligned}
    (1 - \B_{ii}(z))x_i(t) &= v_i(t) + \sum_{k \in \pa{i}}\B_{ik}(z)x_k(t)\\
    \implies x_i(t) &= (1 - \B_{ii}(z))^{-1}\big[v_i(t) + \sum_{k \in \pa{i}}\B_{ik}(z)x_k(t)\big]\\
    &= \wtB_{ii}(z)v_i(t) + \sum_{k \in \pa{i}}\wtB_{ik}(z)x_k(t)
  \end{aligned}
\end{equation*}

% In general, we will write $\wtB$ for arbitrary filters maintaining the
% convention that $\wtB_{ii}$ is a filter that modifies the future of
% $x_i$ from it's past and $\wtB_{ij}$ a filter modifying $x_i$ from the
% past of $x_j$.  We will use the convention that the filter
% $\widetilde{\B}$ will represent simply ``a filter'' and is not
% necessarily the same as the filters serving as edge weights in $\gcg$
% (indeed, $\wtB$ will often be $0$), and moreover, we will allow for
% $\widetilde{\B}$ to potentially change from place to place and even
% from line to line in the same sequence of calculations as it's exact
% properties or specification are not important.

We continue to develop the relationship between $x_i$ and it's family tree.

\begin{proposition}[Ancestor Expansion]
  \label{prop:parent_expanding}
  The component $x_i(t)$ of $x(t)$ can be represented as a sum of
  linear filtering operations acting on the driving noise $v_i$ of
  $x_i$ as well as the parents of $x_i$:

  \begin{equation}
    x_i(t) = \wtB_{ii}(z)v_i(t) + \sum_{k \in \pa{i}}\wtB_{ik}(z)x_k(t).
  \end{equation}

  Moreover, $x_i$ can be expanded in terms of it's ancestors and
  driving noise only:

  \begin{equation}
    \label{eqn:ancestor_expansion}
    x_i(t) = \A_{ii}(z)v_i(t) + \sum_{\substack{k \in \anc{i} \\ k \ne i}}\A_{ik}(z)v_k(t),
  \end{equation}

  where $\A(z) = \sum_{\tau = 0}^\infty A(\tau)$ is the filter from
  the Wold decomposition representation of $x(t)$, equation
  (\ref{eqn:wold}).
\end{proposition}

This statement is ultimately about the sparsity pattern in the Wold
decomposition matrices $A(\tau)$ since
$x_i(t) = \sum_{\tau = 0}^\infty \sum_{j = 1}^n a_{ij}(\tau)v_j(t -
\tau)$.  The proposition states that $\A_{ij}(\tau) = 0$ unless
$j \in \anc{i}$.

% \begin{proof}
%   The formula (\ref{eqn:parent_expansion}) was established in the
%   preceding paragraph and serves as a starting point for the induction
%   hypothesis:

%   \begin{equation}
%     \label{eqn:induction}
%     x_i(t) = \wtB_{ii}(z)v_i(t) + \sum_{k \in \ancn{\ell - 1}{i}\setminus \{i\}}\wtB_{ik}(z)v_k(t) + \sum_{k \in \gpn{\ell}{i} \setminus \{i\}}\wtB_{ik}(z)x_k(t).
%   \end{equation}

%   The base case follows by expanding each $x_k$ in equation \ref{eqn:induction} via equation \ref{eqn:parent_expansion}, but we skip the calculations since they are nearly identical to the following induction step:

%   \begin{align*}
%     x_i(t) &= \wtB_{ii}(z)v_i(t) + \sum_{k \in \ancn{\ell - 1}{i}\setminus \{i\}}\wtB_{ik}(z)v_k(t) + \sum_{k \in \gpn{\ell}{i} \setminus \{i\}}\wtB_{ik}(z)x_k(t)\\
%            &\overset{(a)}{=} \wtB_{ii}(z)v_i(t) + \sum_{k \in \ancn{\ell - 1}{i}\setminus \{i\}}\wtB_{ik}(z)v_k(t) + \sum_{k \in \gpn{\ell}{i} \setminus \{i\}}\wtB_{ik}(z)\big[\wtB_{kk}v_k(t)\\
%     &\ \ \ \  + \wtB_{ki}(z)x_i(t) + \sum_{h \in \pa{k}\setminus \{i\}}\wtB_{kh}(z)x_h(t)\big]\\
%     &\overset{(b)}{=} \wtB_{ii}(z)v_i(t) + \sum_{k \in \ancn{\ell}{i} \setminus \{i\}}\wtB_{ik}(z)v_k(t) + \sum_{k \in \gpn{\ell + 1}{i} \setminus \{i\}}\wtB_{ik}(z)x_k(t).
%   \end{align*}

%   Where equality $(a)$ follows by expanding each $x_k$ with equation \ref{eqn:parent_expansion} and $(b)$ requires the inversion of $\big(1 - \sum_{k \in \gpn{\ell}{i}\setminus\{i\}}\wtB_{ik}(z)\wtB_{ki}(z)\big)$ as well as using the fact that $\ancn{\ell}{i} = \ancn{\ell - 1}{i}\cup\gpn{\ell}{i}$.  We here remind the reader of our earlier established convention that $\wtB$ simply represents the existence of a filter (possibly 0) and can change from place to place.

% The induction terminates after $n$ steps since in a graph with $n$ nodes $\gpn{n + 1}{i} = \emptyset$

% % Where step $(a)$ requires the inversion of $1 - \sum_{k \in \pa{i}}\wtB_{ik}(z)\wtB_{ki}(z)$.

% %   \begin{align*}
% %     x_i(t) &= \wtB_{ii}(z)v_i(t) + \sum_{k \in \pa{i}}\wtB_{ik}(z)x_k(t)\\
% %            &= \wtB_{ii}(z)v_i(t) + \sum_{k \in \pa{i}}\wtB_{ik}(z)\big[\wtB_{kk}v_{k}(t) + \wtB_{ki}(z)x_i(t) + \sum_{h \in \pa{k}\setminus \{i\}}\wtB_{kh}(z)x_h(t) \big]\\
% %            &\overset{(a)}{=} \wtB_{ii}(z)v_i(t) + \sum_{k \in \pa{i}}\wtB_{ik}(z)v_{k}(t) + \sum_{k \in \gpn{2}{i} \setminus \{i\}}\wtB_{ik}(z)x_k(t)
% %   \end{align*}

% % Where step $(a)$ requires the inversion of $1 - \sum_{k \in \pa{i}}\wtB_{ik}(z)\wtB_{ki}(z)$.

% \end{proof}
% \textbf{this is an alternative proof, I think superior to the one above}.

The matrix $\B(z)$ is analagous to a \textit{weighted adjacency
  matrix} for the graph $\gcg$.  And, in the same way that the
$k^{\text{th}}$ power of an adjacency matrix counts the number of
paths of length $k$ between nodes, $(\B(z)^k)_{ij}$ is a filter
isolating the ``action'' of $j$ on $i$ at a time lag of $k$ steps.
Though we only need the sparsity pattern of $\B(z)^k$ for the proof.

We state the following well known theorem as a lemma, proof follows
easily by induction.

\begin{lemma}
  \label{lem:adj_matrix}
  Let $S$ be the transposed adjacency matrix of the Granger-causality
  graph $\gcg$.  Then, $(S^k)_{ij}$ is the number of paths of length
  $k$ from node $j$ to node $i$.  Evidently, if
  $\forall k \in \N,\ (S^k)_{ij} = 0$ then $j \not\in \anc{i}$.
\end{lemma}

We now prove the proposition.

\begin{proof}
  From equation (\ref{eqn:ar_representation}), which we are assuming
  throughout the paper to be invertible, we can write

  \begin{equation*}
    x(t) = (I - \B(z))^{-1} v(t),
  \end{equation*}

  where $(I - \B(z))^{-1} = \A(z)$ due to the uniqueness of
  (\ref{eqn:wold}).  Since $\B(z)$ is stable, that is
  $|\lambda_{\text{max}}(B(z))| < 1$, for every $|z| \le 1$ we have

  \begin{equation}
    \label{eqn:resolvant_inv}
    (I - \B(z))^{-1} = \sum_{k = 0}^\infty \B(z)^k.
  \end{equation}

  Now, let $S$ be a $0-1$ matrix containing the sparsity pattern of
  $\B(z)$.  Then from lemma \ref{lem:adj_matrix} $S$ is the transpose
  of the adjacency matrix for $\gcg$ and hence $(S^k)_{ij}$ is
  non-zero if and only if $j \in \gpn{k}{i}$, and therefore
  $\B(z)^k_{ij} = 0$ if $j \not \in \gpn{k}{i}$.  The Cayley-Hamilton
  theorem can then be applied to represent the infinite sum of equation
  (\ref{eqn:resolvant_inv}) in terms only of \textit{finite} powers of
  $\B(z)$, so the application of lemma \ref{lem:adj_matrix} in the
  limit is not necessary.  Finally, since $\anc{i} = \cup_{k = 1}^n\gpn{k}{i}$
  we see that $\A(z)_{ij}$ is zero if $j \not\in \anc{i}$.

  Thence

  \begin{align*}
    x_i(t) &= (I - \B(z))^{-1}\\
    &= \sum_{j = 1}^n \A_{ij}(z) v_j(t)\\
    &= \A_{ii}(z) v_i(t) + \sum_{\substack{j \in \anc{i} \\ j \ne i}} \A_{ij}(z) v_j(t)\\
  \end{align*}
\end{proof}

Before moving onto our next proposition, we need two lemmas formalizing
some intuitively apparent facts.

\begin{lemma}
  \label{lem:subspace_sum_projection}
  Let $x \in \H$ where $\H$ is a Hilbert space having inner product $\inner{\cdot}{\cdot}$ and let $\{\H_n\}_{n = 0}^\infty$ be a collection of subspaces of $\H$ such that $x \perp \H_0$.  Then

  \begin{equation*}
    \linE{x}{\bigvee_{n = 0}^\infty\H_n} = \linE{x}{\bigvee_{n = 1}^\infty\H_n},
  \end{equation*}

  where $\H_2 \vee \H_2 = \cl \{\alpha + \beta\ |\ \alpha \in \H_1, \beta \in \H_2 \}$ is the closed sum of subspaces.
\end{lemma}
\begin{proof}
  We can apply the Gram-schmidt process to the collection of subspaces to obtain a sequence of orthogonal subspaces $\widetilde{\H}_0, \widetilde{\H}_1, \ldots$ with $\widetilde{\H}_0 = \H_0$, $\widetilde{\H}_n \subseteq \H_n$, and $\bigvee_{n = 0}^\infty\H_n = \bigoplus_{n = 0}^\infty \widetilde{\H}_n$, where $\oplus$ indicates the orthogonal sum of subspaces (which is always closed).  Then, we have an explicit representation of the projection

  \begin{align*}
    \linE{x}{\bigvee_{n = 0}^\infty\H_n} &= \sum_{n = 0}^\infty \inner{x}{x_n} x_n\\
    &\overset{(a)}{=} \sum_{n = 1}^\infty \inner{x}{x_n} x_n,
  \end{align*}

  where $x_n \in \widetilde{\H}_n \subseteq \H_n$ and $(a)$ follows because of our assumption that $x \perp \H_0$.  This establishes that the projection is infact a member of the smaller subspace $\bigvee_{n = 1}^\infty\H_n$, and since the error can only increase by projecting onto this smaller subspace, we have the conclusion.
\end{proof}

The following lemma is reproduced from \cite{lindquist} here for convenience.

\begin{lemma}[\cite{lindquist}]
  \label{lem:conditional_orthogonality_equivalence}
  Consider three closed subspaces of a Hilbert space $\mathcal{A}$, $\mathcal{B}$, $\mathcal{X}$.  The following statements are equivalent

  \begin{enumerate}
    \item{$\mathcal{A} \perp \mathcal{B}\ |\ \mathcal{X}$}
    \item{$\linE{\beta}{\mathcal{A} \vee \mathcal{X}} = \linE{\beta}{\mathcal{X}}\ \forall \beta \in \mathcal{B}$}
    % \item{$\linE{\beta}{\mathcal{A}} = \linE{\linE{\beta}{\mathcal{X}}}{\mathcal{A}}\ \forall \beta \in \mathcal{B}}$}
  \end{enumerate}
\end{lemma}
\begin{proof}
  Proceding from the definition of conditional orthogonality we have

  \begin{align*}
    &\inner{\alpha - \linE{\alpha}{\mathcal{X}}}{\beta - \linE{\beta}{\mathcal{X}}} = 0 \ \forall \alpha \in \mathcal{A}\ \beta \in \mathcal{B}\\
    &{\iff} \inner{\alpha + x - \linE{\alpha + x}{\mathcal{X}}}{\beta - \linE{\beta}{\mathcal{X}}} = 0 \ \forall \alpha \in \mathcal{A}\ \beta \in \mathcal{B}, x \in \mathcal{X}\\
    &\overset{(a)}{\iff} \inner{\alpha + x}{\beta - \linE{\beta}{\mathcal{X}}} = 0 \ \forall \alpha \in \mathcal{A}\ \beta \in \mathcal{B}, x \in \mathcal{X}\\
    &\overset{(b)}{\iff} \linE{\beta - \linE{\beta}{\mathcal{X}}}{\mathcal{A} \vee \mathcal{X}} = 0\ \forall \beta \in \mathcal{B}\\
    &\overset{(c)}{\iff} \linE{\beta}{\mathcal{A} \vee \mathcal{X}} = \linE{\beta}{\mathcal{X}},
  \end{align*}

  where $(a)$ follows since $(\beta - \linE{\beta}{\mathcal{X}}) \perp \mathcal{X}$ and $\linE{\alpha + x}{\mathcal{X}} \in \mathcal{X}$, $(b)$ since $z \perp \mathcal{X} \vee \mathcal{A} \iff \linE{z}{\mathcal{X} \vee \mathcal{A}} = 0$, and $(d)$ since $\linE{z}{\H_2} = \linE{\linE{z}{\H_2}}{\H_1}$ for $\H_2 \subseteq \H_1$.
